---
title: "Mind the gender gap: COVID-19 lockdown effects on gender differences in research"
author: "Iñaki Úcar, Margarita Torre, Antonio Elías Fernández"
output:
  html_document:
    df_print: paged
    number_sections: true
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

<details>
<summary>Functions</summary>
```{r}
get_file <- function(x) {
  dir.create("data", showWarnings=FALSE)
  file <- file.path("data", x)
  if (!file.exists(file)) download.file(
    paste0("https://zenodo.org/record/5142676/files/", x, "?download=1"), file)
  file
}

mutate_cond <- function(.data, condition, ..., envir = parent.frame()) {
  require(dplyr)
  
  condition <- eval(substitute(condition), .data, envir)
  .data[condition, ] <- mutate(.data[condition, ], ...)
  .data
}

get_main <- function(repo, data, n=100) {
  require(dplyr)
  
  subcat <- data %>%
    filter(repository %in% repo) %>%
    group_by(subcategory) %>%
    summarise(weight = n(), .groups="drop") %>%
    arrange(desc(weight)) %>%
    pull(subcategory)
  
  data %>%
    filter(repository %in% repo) %>%
    group_by(id) %>%
    summarise(main = subcat[subcat %in% subcategory][1], .groups="drop") %>%
    group_by(main) %>%
    summarise(N = n(), .groups="drop") %>%
    arrange(desc(N)) %>%
    filter(N >= n) %>%
    pull(main)
}

# performance
model_performance <- function(m) {
  require(ggplot2)
  
  # obs. vs fitted
  df <- data.frame(fitted=fitted(m), response=model.response(model.frame(m)))
  ggplot(df) + aes(fitted, response) +
    geom_point(alpha=0.3) + geom_abline() + geom_smooth(method="lm") +
    ggpmisc::stat_poly_eq(formula=y~x, parse=TRUE)
}
```
</details>

# Data preparation

## Cleaning and filtering

```{r, message=FALSE}
library(dplyr)

articles <- readr::read_csv(get_file("categories.csv")) %>%
  left_join(readr::read_csv(get_file("articles.csv")), by="id") %>%
  rename(subcategory = subcategory_name) %>%
  select(id, repository, category, subcategory, date) %>%
  distinct() %>%
  collect() %>%
  
  # define main categories for repo != arXiv
  mutate_cond(subcategory %in% c("Epidemiology", "Clinical Trials"), repository = "medrxiv") %>%
  mutate_cond(repository == "medrxiv", category = "Health Sciences") %>%
  mutate_cond(repository == "biorxiv", category = "Biology") %>%
  mutate_cond(repository == "psyarxiv", category = "Psychology") %>%
  mutate_cond(repository == "socarxiv", category = "Social Sciences")

articles %>%
  group_by(repository, category) %>%
  summarise(subcategories=length(unique(subcategory)), n=n())
```

```{r, message=FALSE}
articles <- articles %>%
  # mark lockdown period & move time reference to 0
  mutate(year = lubridate::year(date), month = lubridate::month(date)) %>%
  mutate(lockdown = year == 2020 & month > 2) %>%
  mutate(yearmon = (year - min(year)) + (month - 1) / 12) %>%

  # adjustments to Biology and Economics
  mutate_cond(category == "Quantitative Biology", category = "Biology") %>%
  mutate_cond(category == "Quantitative Finance", category = "Economics") %>%
  
  # adjustments to socarxiv
  filter(subcategory != "Social and Behavioral Sciences") %>%
  mutate_cond(repository == "socarxiv" & subcategory %in% c(
    "Science and Technology Studies",
    "Environmental Studies",
    "International and Area Studies",
    "Leisure Studies",
    "Organization Development",
    "Leadership Studies"
    ), subcategory = "Other Social and Behavioral Sciences") %>%
  mutate_cond(repository == "socarxiv" & subcategory %in% c(
    "Library and Information Science",
    "Linguistics"
    ), subcategory = "Arts and Humanities") %>%
  mutate_cond(repository == "socarxiv" & subcategory %in% c(
    "Urban Studies and Planning",
    "Social Statistics"
    ), subcategory = "Sociology") %>%
  mutate_cond(repository == "socarxiv" & subcategory %in% c(
    "Social Work",
    "Legal Studies"
    ), subcategory = "Law") %>%
  mutate_cond(repository == "socarxiv" & subcategory %in% c(
    "Public Affairs, Public Policy and Public Administration"
    ), subcategory = "Political Science") %>%
  mutate_cond(repository == "socarxiv" & subcategory %in% c(
    "Agricultural and Resource Economics"
    ), subcategory = "Economics") %>%
  mutate_cond(id %in% filter(
    ., repository == "socarxiv" & subcategory == "Psychology")$id,
    category = "Psychology") %>%
  filter(subcategory != "Psychology") %>%
  na.omit() %>%
  distinct() %>%
  
  # get just main categories with more than 100 papers for socarxiv & psyarxiv
  filter((!repository %in% c("socarxiv", "psyarxiv")) | subcategory %in% unlist(
    lapply(c("socarxiv", "psyarxiv"), get_main, .)))

articles %>%
  group_by(category) %>%
  summarise(subcategories=length(unique(subcategory)), n=n())
```

```{r, message=FALSE}
text <- readr::read_csv(get_file("text.csv")) %>%
  select(id, title) %>%
  collect() %>%
  mutate(covidpaper = grepl("covid-19", title, ignore.case=TRUE) |
           grepl("sars-cov-2", title, ignore.case=TRUE) |
           grepl("coronavirus", title, ignore.case=TRUE))
```

```{r, message=FALSE}
authors <- readr::read_csv(get_file("authors.csv")) %>%
  distinct() %>%
  collect() %>%
  mutate(across(probability, as.numeric)) %>%
  mutate(across(alphabetical_ordered, as.logical)) %>%
  mutate_cond(is.na(alphabetical_ordered), alphabetical_ordered=TRUE)
```

## Feature engineering

```{r}
merge_info <- function(.data) .data %>%
  left_join(articles, by="id") %>%
  left_join(text[, c("id", "covidpaper")], by="id") %>%
  distinct() %>%
  na.omit() %>%
  mutate_cond(year < 2020, covidpaper = FALSE) %>%
  group_by(subcategory) %>%
  filter(min(year) < 2020) %>%
  ungroup()

# features all
df.id <- authors %>%
  filter(probability >= 0.95 & count >= 10) %>%
  group_by(id) %>%
  summarise(n_male = sum(gender == "male", na.rm=TRUE),
            n_female = sum(gender == "female", na.rm=TRUE),
            n_na = sum(is.na(gender)),
            total = n_male + n_female, .groups="drop") %>%
  merge_info() %>%
  # remove papers with more than 25% of authors with missing gender
  filter(n_na < 0.25 * (n_na + n_male + n_female)) %>%
  # remove observations with less than 30 people per month
  group_by(repository, category, subcategory, covidpaper, lockdown, yearmon) %>%
  filter(sum(total) > 30) %>%
  ungroup()
df.all <- df.id %>%
  group_by(repository, category, subcategory, covidpaper, lockdown, yearmon) %>%
  summarise(p_male = sum(n_male) / sum(total),
            total = sum(total), .groups="drop")

summarise_solo <- function(.data) .data %>%
  group_by(repository, category, subcategory, covidpaper, lockdown, yearmon) %>%
  summarise(n_male = sum(gender == "male", na.rm=TRUE),
            n_female = sum(gender == "female", na.rm=TRUE),
            total = n_male + n_female,
            p_male = n_male / total, .groups="drop")

# features first
df.first <- authors %>%
  filter(probability >= 0.95 & count >= 10) %>%
  filter(!alphabetical_ordered & rank == "first") %>%
  merge_info() %>%
  summarise_solo()

# features last
df.last <- authors %>%
  filter(probability >= 0.95 & count >= 10) %>%
  filter(!alphabetical_ordered & rank == "last") %>%
  merge_info() %>%
  summarise_solo()

# features single
df.single <- authors %>%
  filter(probability >= 0.95 & count >= 10) %>%
  filter(rank == "single") %>%
  merge_info() %>%
  summarise_solo()
```

# Descriptive

## Data after pre-processing

### Number of unique papers

```{r}
articles %>%
  select(id) %>%
  distinct() %>%
  count()
```

### Number of subcategories per category

```{r}
articles %>%
  select(category, subcategory) %>%
  distinct() %>%
  count(category)
```

### Number of papers per category

```{r}
articles %>%
  select(category, id) %>%
  distinct() %>%
  count(category)
```

## Analysis of missings

### Missing authors

```{r}
authors %>%
  select(-id) %>%
  distinct() %>%
  summarise(
    missing = sum(is.na(probability)) / n(),
    discarded = sum(!is.na(probability) & (probability<0.95 | count<10)) / n(),
    total = n()
  )
```

### Missing authors per year

```{r}
articles %>%
  select(id, year) %>%
  distinct() %>%
  right_join(authors) %>%
  group_by(year) %>%
  summarise(
    missing = sum(is.na(probability)) / n(),
    discarded = sum(!is.na(probability) & (probability<0.95 | count<10)) / n(),
    total = n()
  ) %>%
  na.omit()
```

## Preprints considered per year and model

### All authors

```{r}
pp.all <- articles %>%
  select(id, year) %>%
  distinct() %>%
  count(year)
authors %>%
  filter(probability >= 0.95 & count >= 10) %>%
  group_by(id) %>%
  summarise(n_male = sum(gender == "male", na.rm=TRUE),
            n_female = sum(gender == "female", na.rm=TRUE),
            n_na = sum(is.na(gender)),
            total = n_male + n_female, .groups="drop") %>%
  merge_info() %>%
  # remove papers with more than 25% of authors with missing gender
  filter(n_na < 0.25 * (n_na + n_male + n_female)) %>%
  select(id, year) %>%
  distinct() %>%
  count(year) %>%
  mutate(prop = n / pp.all$n)
```

### First authors

```{r}
pp.first <- authors %>%
  filter(!alphabetical_ordered & rank == "first") %>%
  merge_info() %>%
  select(id, year) %>%
  distinct() %>%
  count(year)
authors %>%
  filter(probability >= 0.95 & count >= 10) %>%
  filter(!alphabetical_ordered & rank == "first") %>%
  merge_info() %>%
  select(id, year) %>%
  distinct() %>%
  count(year) %>%
  mutate(prop = n / pp.first$n)
```

### Last authors

```{r}
pp.last <- authors %>%
  filter(!alphabetical_ordered & rank == "last") %>%
  merge_info() %>%
  select(id, year) %>%
  distinct() %>%
  count(year)
authors %>%
  filter(probability >= 0.95 & count >= 10) %>%
  filter(!alphabetical_ordered & rank == "last") %>%
  merge_info() %>%
  select(id, year) %>%
  distinct() %>%
  count(year) %>%
  mutate(prop = n / pp.last$n)
```

### Single authors

```{r}
pp.single <- authors %>%
  filter(rank == "single") %>%
  merge_info() %>%
  select(id, year) %>%
  distinct() %>%
  count(year)
authors %>%
  filter(probability >= 0.95 & count >= 10) %>%
  filter(rank == "single") %>%
  merge_info() %>%
  select(id, year) %>%
  distinct() %>%
  count(year) %>%
  mutate(prop = n / pp.single$n)
```

## Evolution per category over time

### Number of papers per month

```{r}
library(ggplot2)
theme_set(theme_bw())

df.plot.papers <- df.id %>%
  select(yearmon, category, id) %>%
  distinct() %>%
  count(yearmon, category)
df.plot.papers.all <- df.plot.papers %>%
  group_by(yearmon) %>%
  summarise(n = sum(n), .groups="drop") %>%
  mutate(category="(all)")

ggplot(df.plot.papers) +
  aes(yearmon+2017, n) + facet_wrap(~category, scales="free_y") +
  geom_col(data=df.plot.papers.all) +
  geom_smooth(data=df.plot.papers.all) +
  geom_col() + geom_smooth() +
  zoo::scale_x_yearmon(format="%Y") +
  labs(y="Number of papers per month", x=NULL)
```

### Number of authors per month

```{r}
df.plot.authors <- df.id %>%
  select(yearmon, category, id, total) %>%
  distinct() %>%
  group_by(yearmon, category) %>%
  summarise(n = sum(total), .groups="drop")
df.plot.authors.all <- df.plot.authors %>%
  group_by(yearmon) %>%
  summarise(n = sum(n), .groups="drop") %>%
  mutate(category="(all)")

ggplot(df.plot.authors) +
  aes(yearmon+2017, n) + facet_wrap(~category, scales="free_y") +
  geom_col(data=df.plot.authors.all) +
  geom_smooth(data=df.plot.authors.all) +
  geom_col() + geom_smooth() +
  zoo::scale_x_yearmon(format="%Y") +
  labs(y="Number of authors per month", x=NULL)
```

### Proportion of males per month

```{r}
df.plot.p_male <- df.id %>%
  select(yearmon, category, id, n_male, total) %>%
  distinct() %>%
  group_by(yearmon, category) %>%
  summarise(p_male = sum(n_male) / sum(total), .groups="drop")
df.plot.p_male.all <- df.id %>%
  group_by(yearmon) %>%
  summarise(p_male = sum(n_male) / sum(total), .groups="drop") %>%
  mutate(category="(all)")

ggplot(df.plot.p_male) +
  aes(yearmon+2017, p_male) + facet_wrap(~category) +
  geom_point(data=df.plot.p_male.all) +
  geom_smooth(method="gam", data=df.plot.p_male.all) +
  geom_point() + geom_smooth(method="gam") +
  zoo::scale_x_yearmon(format="%Y") +
  labs(y="Proportion of male authors", x=NULL)
```

# Modelling

## Simple model

```{r}
model.glm <- p_male ~ yearmon + lockdown + covidpaper
fit.glm <- glm(model.glm, df.all, family=binomial, weights=total)
summary(fit.glm)
```

```{r}
model_performance(fit.glm)
performance::check_model(fit.glm)
```

## Hierarchical model

```{r}
library(lme4)

model <- p_male ~ yearmon + lockdown + covidpaper + (1 | category/subcategory)
fit.all <- glmer(model, df.all, family=binomial, weights=total)
summary(fit.all)
```

```{r}
model_performance(fit.all)
performance::check_model(fit.all)
```
```{r}
df.re.category <- sjPlot::get_model_data(fit.all, "re")[[2]] %>%
  mutate(group = ifelse(conf.low < 1 & conf.high > 1, NA, group)) %>%
  mutate(term = forcats::fct_reorder(term, estimate))

ggplot(df.re.category) +
  aes(estimate, term, xmin=conf.low, xmax=conf.high, color=group) +
  geom_vline(xintercept=1, color="black", linetype="dashed") +
  geom_errorbarh(height=0, size=1) + geom_point(size=2) +
  scale_x_log10(breaks=c(0.2, 0.5, 1, 2, 5)) +
  scale_color_manual(
    breaks=c("pos", "neg"), values=c("blue", "red"),
    na.value=grDevices::adjustcolor("grey50", 0.5)) +
  labs(y=NULL, x="Odds ratio", title="Random effect: category") +
  theme(legend.position="none")
```

```{r, fig.asp=4}
df.re.subcategory <- sjPlot::get_model_data(fit.all, "re")[[1]] %>%
  mutate(group = ifelse(conf.low < 1 & conf.high > 1, NA, group)) %>%
  mutate(facet = sapply(strsplit(as.character(term), ":"), tail, 1)) %>%
  mutate(term = forcats::fct_reorder(term, estimate))

ggplot(df.re.subcategory) +
  aes(estimate, term, xmin=conf.low, xmax=conf.high, color=group) +
  facet_grid("facet", scales="free_y", space="free_y") +
  geom_vline(xintercept=1, color="black", linetype="dashed") +
  geom_errorbarh(height=0, size=1) + geom_point(size=2) +
  scale_x_log10(breaks=c(0.2, 0.5, 1, 2, 5)) +
  scale_color_manual(
    breaks=c("pos", "neg"), values=c("blue", "red"),
    na.value=grDevices::adjustcolor("grey50", 0.5)) +
  labs(y=NULL, x="Odds ratio", title="Random effect: subcategory") +
  theme(legend.position="none")
```

## All together

```{r}
fit <- list(
  glm = fit.glm,
  all = fit.all,
  first = glmer(model, df.first, family=binomial, weights=total),
  last = glmer(model, df.last, family=binomial, weights=total),
  single = glmer(model, df.single, family=binomial, weights=total)
)

df.terms <- lapply(fit, sjPlot::get_model_data, "est") %>%
  bind_rows(.id="fit") %>%
  mutate(type = ifelse(fit=="glm", "GLM", "GLMM")) %>%
  mutate(fit = ifelse(fit=="glm", "all", fit)) %>%
  mutate(fit = forcats::fct_inorder(fit)) %>%
  mutate(type = ifelse(p.value >= 0.05, NA, type))

levels(df.terms$term) <- c("COVID paper", "lockdown", "year")
```

```{r, results='asis'}
stargazer::stargazer(fit, type="html")
```

```{r, fig.asp=1}
ggplot(df.terms) +
  aes(estimate, term, xmin=conf.low, xmax=conf.high, color=type) +
  facet_grid("fit", scales="free_y", space="free_y") +
  geom_vline(xintercept=1, color="black", linetype="dashed") +
  geom_errorbarh(height=0, size=1) + geom_point(size=2) +
  ggrepel::geom_text_repel(
    aes(label=p.label), nudge_y=0.3, nudge_x=0.01, segment.size=0, show.legend=FALSE) +
  scale_x_log10(breaks=c(0.2, 0.5, 1, 2, 5)) +
  scale_color_manual(
    breaks=c("GLM", "GLMM"), values=c("red", "black"),
    na.value=grDevices::adjustcolor("grey50", 0.5)) +
  labs(y=NULL, x="Odds ratio", color="Model") +
  theme(legend.position=c(.99, .99), legend.justification=c(1, 1)) +
  theme(legend.background=element_rect(fill="#ffffff88"))
```
