---
title: 'COVID dataset: load the data'
author: "Antonio Elías"
date: "06/07/2020"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Install and load the library RSQLite

```{r}
#install.packages("RSQLite")
library(RSQLite)
```


# 2. Load the data set

## 2.1 Create connection to the db
```{r}
conn_table_CONCIERGE <- dbConnect(RSQLite::SQLite(), "database/dataRxiv.db")
```

## 2.2 See list of tables
```{r}
dbListTables(conn_table_CONCIERGE)
```

We have four tables ARTICLES, AUTHORS, CATEGORIES AND TEXT.

## 2.3 Get the tables as data.frames

Table ARTICLES includes:

  - **id**: article identifier.
  - **repository**: repository name.
  - **date**: first date of publication (YEAR-MONTH-DAY).

```{r}
articles <- dbGetQuery(conn_table_CONCIERGE, "SELECT * FROM ARTICLES")
head(articles)
```

Table AUTHORS includes:

  - **id**: article identifier.
  - **full_name**: full name.
  - **given_name**: first word that appears at full_name.
  - **alphabetical_ordered**: if the full names of the article are alphabetically ordered (TRUE) or (FALSE).
  - **position**: author position 1, 2, 3...
  - **rank**: first authors, middle author, last author or single author.
  - **name**: named used for "genderizing" the author.
  - **gender**: predicted gender (male or female).
  - **probability**: number of times that the number appears as gender X divided by the number of times the name appears in the genderize data set.
  - **count**: number of times that the name appears in the genderize data set.

```{r}
authors <- dbGetQuery(conn_table_CONCIERGE, "SELECT * FROM AUTHORS")
head(authors)
```

Table CATEGORIES includes:

  - **id**: article identifier.
  - **category**: first level of category (arxiv).
  - **subcategory_code**: code given in arxiv.
  - **subcategory_name**: same as code for arxiv (but understandable). For the other repositories this is the subject.
  - **primary**: if it is the primary category (for arxiv takes 0 and 1) for other repositories is always 1.

```{r}
categories <- dbGetQuery(conn_table_CONCIERGE, "SELECT * FROM CATEGORIES")
head(categories)
```

Table TEXT includes:

  - **id**: article identifier.
  - **title**: article title.
  - **abstract**: article abstract.

```{r}
text <- dbGetQuery(conn_table_CONCIERGE, "SELECT * FROM TEXT")
head(text)
```

Changes in articles, authors, categories or text do not modify the main dataRxiv.db.

## 2.4 Disconnect to the database
Disconnect the database.

```{r}
dbDisconnect(conn_table_CONCIERGE)
```
